---
title: "Data Wrangling Workbook"
author: "PSYC417"
format: html
editor: visual
---

## Installing & loading appropriate libraries (tidyverse)

```{r}
# Recall: You will need to comment out (using #) the install.packages("tidyverse") line of code before you render.
# You only need to run the install.packages("tidyverse") line of code once! You don't need to rerun it each time you open a new Quarto document.
install.packages("tidyverse")

# You must run the library("tidyverse") each time you open R.
library("tidyverse")
```

# Let's take a look at the dataframe we will be working with

```{r}
starwars <- as.data.frame(starwars)
```

# A few basics notes

```{r}
# There is a difference between creating an output and creating a new dataframe

# Creating an output
starwars %>%
  filter(species == "Droid")

# Creating a new dataframe 
starwars.droids <- starwars %>%
  filter(species == "Droid")

# Why does this matter? If you do not create a new dataframe, you cannot use the manipulated data later on! 
```

## Functions within Tidyverse

# filter() - 1 condition

```{r}
# I want to index data to only include humans
# species = Human


```

# filter() - 2 conditions

```{r}
# I want to index data to only include humans who are from the Naboo planet
# species = Human
# homeworld = Naboo


```

# filter() - 3+ conditions

```{r}
# I want to index data to only include humans, who are from the Naboo planet, and have brown hair
# species = Human
# homeworld = Naboo
# hair_color = brown



# You can add more conditions by using more commas

```

# filter() - remove missing data (for a specific variable)

```{r}
# In the previous output, we can see there's some missing data. Let's remove that! 
# We can remove missing data from specific variables, in this example, we will remove it from the mass variable


# Now, you have an output that includes humans, who are from the Naboo planet, have brown hair with all missing data removed

```

# arrange() - ascending order (default)

```{r}
# I want to arrange my data from characters with the smallest mass to the largest mass

```

# arrange() - descending order

```{r}
# I want to arrange my data from characters with the largest mass to the smallest mass

```

# slice()

```{r}
# I want the 30-40th rows of the dataframe

```

# slice_head()

```{r}
# I want the top 12 rows of the dataframe

```

# slice_tail()

```{r}
# I want the bottom 11 rows of the dataframe

```

# slice_sample() - specific number

```{r}
# I want 14 random rows from the dataframe

```

# slice_sample() - proportion

```{r}
# I want a random 10% of rows from the dataframe

```

# slice_min()

```{r}
# I want the 3 rows with the lowest height (the 3 shortest characters)

```

# slice_max()

```{r}
# I want the 4 rows with the highest height (the 3 tallest characters)

```

# select() - specifying the columns we want

```{r}
# I want a dataframe that only includes the variables: name, homeworld, and species

```

# select() - removing the columns we don't want

```{r}
# I want a dataframe that only includes all the variables except films

```

# select(starts_with())

```{r}
# I want a dataframe that only includes the variables that start with the letter "s"

```

# select(ends_with())

```{r}
# I want a dataframe that only includes the variables that start with the letter "r"

```

# select(contains())

```{r}
# I want a dataframe that only includes the variables that contain the world "color" in them

```

# select() - change the order of the first variable, then keep all other variables in the same order

```{r}
# I want a dataframe that has the homeworld variable listed first

```

# mutate() - 1 new variable

```{r}
# I want to add a variable called 'combined' that adds the character's height and mass together (height + mass)

```

# mutate() - 2 new variables

```{r}
# I want to add a variable called 'combined' that adds the character's height and mass together (height + mass) and another variable called 'divided' that divides their height by mass (height/mass)

```

# summarise()

```{r}
# This function doesn't work well with our data, so I have created a new dataframe below for the sake of the example 

# Let's say we have the following dataframe (go ahead and take a look):
data.example <- data.frame(subjectID = 1:5,
                           wmc.score = c(23,42,12,24,18),
                           ltm.score = c(18, 21, 22, 19, 19))

# I want to get the mean wmc.score called "mean.wmc" & the mean ltm.score called "mean.ltm"


# I want to get the mean wmc.score called "mean.wmc" & the standard deviation of wmc.score called "sd.wmc"

```

# group_by()

```{r}
# group_by() is usually combined with another function

# I want the mean height of each species 
# So, we will group by species and summarise the mean height

```

# summarise() - remove missing data (for a specific variable)

```{r}
# In the previous output, we can see there's some missing data. Let's remove that! 
# We can remove missing data from the new summarised score we created (mean.height)

```

# Combining multiple actions together

```{r}
# For humans only, I want the mean height based on eye color. But, I only want to know what the the two highest mean heights from that is

```

## For pivot_longer() we will be using the Seatbelts dataframe.

```{r}
# Let's take a look at our dataframe
Seatbelts <- as.data.frame(Seatbelts)
```

# pivot_longer()

```{r}
# I want the dataframe to have 1 column that specifies location in the car (front and rear) called "Location" and 1 column that indicates the number of deaths for passengers in that location called "Deaths"

```

## For pivot_wider(), we will be using the ChickWeight dataframe.

```{r}
#Let's take a look at our dataframe
ChickWeight <- as.data.frame(ChickWeight)
```

# pivot_wider()

```{r}
# I want the dataframe to show one row per chick where there are separate columns for each day

```
